Geometric variational problems

### 2020-11-03 NSF2020/Description/specificaim2_ver2.tex
working on proposal

\subsection{Specific Aim 2: Efficient high-order numerical algorithm for
large-scale simulation of the HAP model}
\label{subsec:specific_aim_2}
% ----------------------------------------------------------------------
\subsubsection{Overview of the potential theory and boundary integral
equation methods} 
The HAP model requires solving the exterior Dirichlet problem of the
screened Laplace equation and the mobility problem of the Stokes flow at
each time step. If discretizing these equations with stencil-based
numerical methods, the computational domain must be truncated, the
volume must be discretized, and artificial boundary conditions must be
imposed. The artificial boundary conditions introduces an additional
source of error, and discretizing the volume, particularly in three
dimensions, results in an excessively large linear system. Moreover, it
is difficult to obtain high-order discretizations when the boundary
surfaces are irregular. Since we are solving constant-coefficient
elliptic PDEs, we apply boundary integral equation (BIE) that remove the
aforementioned obstacles. BIEs represent the solution as a layer
potential with an unknown density function. Then, the correct density
function is found that matching the layer potential with the prescribed
boundary conditions. By construction, layer potentials satisfy the
underlying PDE and the far-field conditions.

Discretizations of BIEs have several advantages when compared to
stencil-based discretizations. First, the dimension of the problem is
reduced by one since the unknown density function is defined on the
boundary of the geometry. Second, by carefully selecting each layer
potential, the resulting linear system is well-conditioned for any
resolution and iterative methods such as GMRES require a
mesh-independent number of iterations to converge. Third, by using
appropriate quadrature methods, the method converges with spectral or
very high-order accuracy.

Challenges are dense matrices, preconditioners, and near-singular
integration

Potential theory and boundary integral equation (BIE) methods remove the
aforementioned obstacles in a very elegant way. The starting point is
that for constant-coefficient PDEs the fundamental solutions, i.e., the
Green's functions, are readily available. This allows us to represent
the solution via the so-called layer potentials that are the convolution
integrals of a kernel and an unknown density only on the boundary. The
kernel is a linear combination of the Green's function and its
derivatives, which ensures that the representation satifies the
underlying PDE and the condition at infinity automatically. And the
boundary condition at the material interfaces together with the jump
relation of the layer potential leads to a boundary integral equation
for the unknown density.

The discretization of the integral operator often leads to a dense
matrix, while finite difference/finite element methods result in sparse
matrices, making matrix-vector product much cheaper to compute.
However,since the invention of the original fast multipole method
(FMM)\cite{fmm5}, there have been many fast
algorithms~\cite{fmm1,fmm2,fmm3,fmm4,fmm6,fmm7,fmm8} that reduce the
cost of matrix-vector product to linear ($O(N)$) or quasilinear
complexity ($O(N \log^kN)$) for dense matrices resulting from the BIE
discretization. Fast direct solvers have also been developed
recently~\cite{fds1,fds2,fds3,fds4,fds5,fds6,fds7,fds8,ho2016cpam2,ho2016cpam1,minden2016,minden2017siammms}.
The development of these fast algorithms have removed a huge hurdle for
the use of boundary integral equations, making the BIE method the method
of choice for constant-coefficient PDEs with linear boundary conditions.
For the exterior Dirichlet problem of the screened Laplace equation, the
SKIE formulation in our previous work~\cite{Fu2018_SIAM} can be readily
extened to the 3D case. For the mobility problem of the Stokes flow in
three dimensions that takes the hydrodynamic effect into account, an
SKIE formulation can be found in~\cite{manasthesis}.

% ----------------------------------------------------------------------
\subsubsection{High-order discretization of surface integrals in three
dimensions}
The practical application of integral equation methods requires the
accurate evaluation of boundary integrals with singular, weakly singular
or nearly singular kernels. There are many numerical quadrature schemes
for dealing with line integrals in two
dimensions~\cite{alpert,kapur,sidi,duffy,bruno1,bruno2,davis_1984,graglia_2008,hackbusch_sauter_1994,
jarvenpaa_2003,khayat_2005,kress_boundary_1991,schwab_1992,
ying_2006,beale1,beale2,goodman_1990, haroldson_1998,
lowengrub_1993,schwab_1992,ggq1,ggq2,ggq3,helsing_2008a,helsing_integral_2009,helsing_tutorial_2012}
and some of them are extremely efficient and can achieve arbitrary high
order.

The high-order quadrature rules for the evaluation of surface integrals
in three dimensions, however, are much less developed than the line
integrals in two dimensions. For example, there are no Gaussian
quadratures for integraing polynomials on a flat triangle, even though
efficient quadratures~\cite{xiao2010cma,vioreanu2014} have been
developed recently for such purpose. For weakly singular or singular
integrals,~\cite{bremer2012jcp,bremer2013jcp} constructed high-order
quadratures for surface integrals on a general triangle,
while~\cite{gimbutas2013sisc} presented a fast algorithm for integrating
$1/r$-type singular integrals for surfaces that are homeomorphic to a
sphere. We plan to study the so-called quadrature by expansion (QBX)
scheme~\cite{klockner2013jcp,qbx2} for the evaluation of both singular
and near-singular surface integrals encountered in the discretization of
BIEs in three dimensions. Conceptually, the idea of the QBX to evaluate
singular, hypersingular and near singular integrals on smooth surfaces
is straightforward. That is, the surface is discretized into smooth
triangles and smooth high-order quadratures are applied to evaluate the
expansion coefficients on all source triangles with the QBX expansion
center placed at a point off the surface. One may then form a suitable
expansion (for example, a Taylor expansion) around that center and
evaluate this expansion back at the target point on the surface (or
close to the surface in the near singular case). Compared to the
competing aforementioned quadrature schemes, the QBX quadrature is
attractive because it offers a clear path for being extended to:
\textbf{(1)} handle any singularity, including hypersingular operators,
\textbf{(2)} be usable with any high-order surface discretization,
\textbf{(3)} generate well-conditioned discrete operators to which
iterative methods such as GMRES~\cite{gmres} can be applied in a
black-box fashion, \textbf{(4)} integrate well with fast algorithms such
as the FMM.

In practice, there are still many issues that need to be resolved. For
example, there are now many variants of QBX including global and local
QBX~\cite{klockner2013jcp,rachh2017jcp}, the target-specific
QBX~\cite{siegel2018jcp}, kernel-independent QBX~\cite{abtin2018bit},
and quadrature by two expansions~\cite{ding2019arxiv}. The coupling of
the QBX and the FMM may also lead to certain instability issues which
may require some changes in the fast multipole
method~\cite{wala2018jcp}. Similar to other quadrature methods, there
have been extensive study on the QBX methods in two dimensions, while
its three dimension
version~\cite{wala2019jcp,af2018sisc,siegel2018jcp,wala2019arxiv} has
not been fully studied and the implementation is even more scarce. We
plan to investigate the accuracy and the convergence order of the
various QBX schemes mentioned above, its coupling with the FMM, parallel
implementation issues for large-scale problems, and the application to
our target problems.

% ----------------------------------------------------------------------
\subsubsection{Other numerical issues}
Another outstanding numerical issue pertaining to simulating the
self-assembly of amphiphilic particles (such as lipid macromolecles) in
a viscous solvent is the collision between amphiphilic particles. The
hydrophobic attraction potential drives the amphiphilic particles to
move towards each other so to minimize exposure to the solvent. Under
such attraction force the fluid is being squeezed out as particles move
toward each other, and the amphiphilic particles can come into physical
contact in finite time. Such particle collisions in a dense rigid body
suspension is a great challenge and can be a bottleneck in large-scale
simulations. In our previous work a Leonard-Jones potential is used to
keep the particles from overlapping~\cite{Fu2018_SIAM}. Such steep
steric interaction at short ranges introduces great numerical stiffness
that limit the time step size.

@@

>>>>

\subsection{Specific Aim 2: Efficient high-order numerical algorithm for
large-scale simulation of the HAP model}
\label{subsec:specific_aim_2}
% ----------------------------------------------------------------------
\subsubsection{Overview of the potential theory and boundary integral
equation methods} 
The HAP model requires solving the exterior Dirichlet problem of the
screened Laplace equation and the mobility problem of the Stokes flow at
each time step. If discretizing these equations with stencil-based
numerical methods, the computational domain must be truncated, the
volume must be discretized, and artificial boundary conditions must be
imposed. The artificial boundary conditions introduces an additional
source of error, and discretizing the volume, particularly in three
dimensions, results in an excessively large linear system. Moreover, it
is difficult to obtain high-order discretizations when the boundary
surfaces are irregular. Since we are solving constant-coefficient
elliptic PDEs, we apply boundary integral equation (BIE) that remove the
aforementioned obstacles. BIEs represent the solution as a layer
potential with an unknown density function. Then, the correct density
function is found that matching the layer potential with the prescribed
boundary conditions. By construction, layer potentials satisfy the
underlying PDE and the far-field conditions. For the exterior Dirichlet
problem of the screened Laplace equation, the BIE formulation in our
previous work~\cite{Fu2018_SIAM} can be readily extended to three
dimensions. For the mobility problem of the Stokes flow in three
dimensions that takes the hydrodynamic effect into account, an
well-conditioned BIE is found in~\cite{manasthesis}.

Discretizations of BIEs have several advantages when compared to
stencil-based discretizations. First, the dimension of the problem is
reduced by one since the unknown density function is defined on the
boundary of the geometry. Second, by carefully selecting each layer
potential, the resulting linear system is well-conditioned for any
resolution and iterative methods such as GMRES require a
mesh-independent number of iterations to converge. Third, by using
appropriate quadrature methods, the method converges with spectral or
very high-order accuracy. The most significant challenges include
solving dense linear systems, developing preconditioning strategies, and
developing quadrature methods for integrands that are nearly singular.

By choosing BIEs that use a mesh-independent number of GMRES iterations,
the overall cost of a solver is proportional to the complexity of the
matrix-vector multiplication. The matrix-vector multiplication will be
accelerated to optimal or near-optimal complexity using the fast
multipole method (FMM)~\cite{fmm5} and its extensions~\cite{fmm1, fmm2,
fmm3, fmm4, fmm6, fmm7, fmm8}. Alternatively, fast direct solvers can be
used~\cite{fds1, fds2, fds3, fds4, fds5, fds6, fds7, fds8, ho2016cpam2,
ho2016cpam1, minden2016, minden2017siammms}, but since direct solvers
often require a large amount of initial overhead, they are less
applicable to problems with moving geometries such as ours.
Alternatively, fast direct solvers can be used to precondition the
linear system as described in Section~\ref{subsec:NumericalIssues}.

% ----------------------------------------------------------------------
\subsubsection{Nearly touching bodies}
The trapezoid rule is the workhorse for BIEs since it achieves spectral
accuracy for smooth periodic functions~\cite{tre-wei2014}. However,
nearly touching bodies is ubiquitous in self-assembly of amphiphilic
particles, and such near-contact results in  nearly-singular integrands.
Quadrature methods to address such integrands has received a lot of
attention in two and three dimensions~\cite{alpert, kapur, sidi, duffy,
bruno1, bruno2, davis_1984, graglia_2008, hackbusch_sauter_1994,
jarvenpaa_2003, khayat_2005, schwab_1992, ying_2006, beale1, beale2,
goodman_1990, haroldson_1998, lowengrub_1993, schwab_1992, ggq1, ggq2,
ggq3, helsing_2008a, helsing_integral_2009, helsing_tutorial_2012,
klockner2013jcp, qbx2, wala2019jcp, af2018sisc, siegel2018jcp,
rachh2017jcp, ding2019arxiv, bar2014}. For two-dimensional problems, an
elegant quadrature method, which we will call a {\em barycentric
quadrature rule}, is a mildly intrusive modification of the trapezoid
rule relies on analytic function theory~\cite{ioa-pap-per1991}. Since
the Stokes equations can be cast in terms of analytic functions, the
barycentric-quadrature rule has recently been
applied~\cite{bar-wu-vee2015, chi-moo-qua2020}. The PIs plan to extend
this quadrature rule so that it can be applied to the screened Laplace
equation.

Another outstanding numerical issue pertaining to simulating the
self-assembly of amphiphilic particles (such as lipid macromolecles) in
a viscous solvent is the collision between amphiphilic particles. The
hydrophobic attraction potential drives the amphiphilic particles to
move towards each other so to minimize exposure to the solvent. Under
such attraction force the fluid is being squeezed out as particles move
toward each other, and the amphiphilic particles can come into physical
contact in finite time. Such particle collisions in a dense rigid body
suspension is a great challenge and can be a bottleneck in large-scale
simulations. In our previous work a Leonard-Jones potential is used to
keep the particles from overlapping~\cite{Fu2018_SIAM}. Such steep
steric interaction at short ranges introduces great numerical stiffness
that limit the time step size.

@@

% ----------------------------------------------------------------------
\subsubsection{Fluctuating hydrodynamics}
Using DLP formulation for fluctuating hydrodynamics so that we have a
second-kind BIE

% ----------------------------------------------------------------------
\subsubsection{Other numerical issues}
\label{subsec:NumericalIssues}
Preconditioning
Adaptive high-order time stepping

%Potential theory and boundary integral equation (BIE) methods remove the
%aforementioned obstacles in a very elegant way. The starting point is
%that for constant-coefficient PDEs the fundamental solutions, i.e., the
%Green's functions, are readily available. This allows us to represent
%the solution via the so-called layer potentials that are the convolution
%integrals of a kernel and an unknown density only on the boundary. The
%kernel is a linear combination of the Green's function and its
%derivatives, which ensures that the representation satifies the
%underlying PDE and the condition at infinity automatically. And the
%boundary condition at the material interfaces together with the jump
%relation of the layer potential leads to a boundary integral equation
%for the unknown density.

%The discretization of the integral operator often leads to a dense
%matrix, while finite difference/finite element methods result in sparse
%matrices, making matrix-vector product much cheaper to compute.
%However,since the invention of the original fast multipole method
%(FMM)\cite{fmm5}, there have been many fast
%algorithms~\cite{fmm1,fmm2,fmm3,fmm4,fmm6,fmm7,fmm8} that reduce the
%cost of matrix-vector product to linear ($O(N)$) or quasilinear
%complexity ($O(N \log^kN)$) for dense matrices resulting from the BIE
%discretization. Fast direct solvers have also been developed
%recently~\cite{fds1,fds2,fds3,fds4,fds5,fds6,fds7,fds8,ho2016cpam2,ho2016cpam1,minden2016,minden2017siammms}.
%The development of these fast algorithms have removed a huge hurdle for
%the use of boundary integral equations, making the BIE method the
%method of choice for constant-coefficient PDEs with linear boundary
%conditions.  For the exterior Dirichlet problem of the screened Laplace
%equation, the SKIE formulation in our previous work~\cite{Fu2018_SIAM}
%can be readily extened to the 3D case. For the mobility problem of the
%Stokes flow in three dimensions that takes the hydrodynamic effect into
%account, an SKIE formulation can be found in~\cite{manasthesis}.

%% ----------------------------------------------------------------------
%\subsubsection{High-order discretization of surface integrals in three
%dimensions}
%The practical application of integral equation methods requires the
%accurate evaluation of boundary integrals with singular, weakly singular
%or nearly singular kernels. There are many numerical quadrature schemes
%for dealing with line integrals in two
%dimensions~\cite{alpert,kapur,sidi,duffy,bruno1,bruno2,davis_1984,graglia_2008,hackbusch_sauter_1994,
%jarvenpaa_2003,khayat_2005,kress_boundary_1991,schwab_1992,
%ying_2006,beale1,beale2,goodman_1990, haroldson_1998,
%lowengrub_1993,schwab_1992,ggq1,ggq2,ggq3,helsing_2008a,helsing_integral_2009,helsing_tutorial_2012}
%and some of them are extremely efficient and can achieve arbitrary high
%order.
%
%The high-order quadrature rules for the evaluation of surface integrals
%in three dimensions, however, are much less developed than the line
%integrals in two dimensions. For example, there are no Gaussian
%quadratures for integraing polynomials on a flat triangle, even though
%efficient quadratures~\cite{xiao2010cma,vioreanu2014} have been
%developed recently for such purpose. For weakly singular or singular
%integrals,~\cite{bremer2012jcp,bremer2013jcp} constructed high-order
%quadratures for surface integrals on a general triangle,
%while~\cite{gimbutas2013sisc} presented a fast algorithm for integrating
%$1/r$-type singular integrals for surfaces that are homeomorphic to a
%sphere. We plan to study the so-called quadrature by expansion (QBX)
%scheme~\cite{klockner2013jcp,qbx2} for the evaluation of both singular
%and near-singular surface integrals encountered in the discretization of
%BIEs in three dimensions. Conceptually, the idea of the QBX to evaluate
%singular, hypersingular and near singular integrals on smooth surfaces
%is straightforward. That is, the surface is discretized into smooth
%triangles and smooth high-order quadratures are applied to evaluate the
%expansion coefficients on all source triangles with the QBX expansion
%center placed at a point off the surface. One may then form a suitable
%expansion (for example, a Taylor expansion) around that center and
%evaluate this expansion back at the target point on the surface (or
%close to the surface in the near singular case). Compared to the
%competing aforementioned quadrature schemes, the QBX quadrature is
%attractive because it offers a clear path for being extended to:
%\textbf{(1)} handle any singularity, including hypersingular operators,
%\textbf{(2)} be usable with any high-order surface discretization,
%\textbf{(3)} generate well-conditioned discrete operators to which
%iterative methods such as GMRES~\cite{gmres} can be applied in a
%black-box fashion, \textbf{(4)} integrate well with fast algorithms such
%as the FMM. 
%
%In practice, there are still many issues that need to be resolved. For
%example, there are now many variants of QBX including global and local
%QBX~\cite{klockner2013jcp,rachh2017jcp}, the target-specific
%QBX~\cite{siegel2018jcp}, kernel-independent QBX~\cite{abtin2018bit},
%and quadrature by two expansions~\cite{ding2019arxiv}. The coupling of
%the QBX and the FMM may also lead to certain instability issues which
%may require some changes in the fast multipole
%method~\cite{wala2018jcp}. Similar to other quadrature methods, there
%have been extensive study on the QBX methods in two dimensions, while
%its three dimension
%version~\cite{wala2019jcp,af2018sisc,siegel2018jcp,wala2019arxiv} has
%not been fully studied and the implementation is even more scarce. We
%plan to investigate the accuracy and the convergence order of the
%various QBX schemes mentioned above, its coupling with the FMM, parallel
%implementation issues for large-scale problems, and the application to
%our target problems.

%% ----------------------------------------------------------------------
%\subsubsection{Other numerical issues}
%Another outstanding numerical issue pertaining to simulating the
%self-assembly of amphiphilic particles (such as lipid macromolecles) in
%a viscous solvent is the collision between amphiphilic particles. The
%hydrophobic attraction potential drives the amphiphilic particles to
%move towards each other so to minimize exposure to the solvent. Under
%such attraction force the fluid is being squeezed out as particles move
%toward each other, and the amphiphilic particles can come into physical
%contact in finite time. Such particle collisions in a dense rigid body
%suspension is a great challenge and can be a bottleneck in large-scale
%simulations. In our previous work a Leonard-Jones potential is used to
%keep the particles from overlapping~\cite{Fu2018_SIAM}. Such steep
%steric interaction at short ranges introduces great numerical stiffness
%that limit the time step size. 
%
%Yan {\it et al.} proposed a collision-resolution algorithm within the
%framework of boundary integral methods to resolve the non-smooth
%many-body dynamics due to collisions by formulating the particle
%collision process as a linear complementarity problem with geometric
%`non-overlapping' constraint~\cite{Yan2019}. We propose to treat the
%particle collisions based on Yan {\it et al.}'s work after the first two
%goals (see above) are achieved. A potential complication in combining
%the algorithms for particle collision into our QBX integral formulation
%is the computational cost for the QBX to refine meshes on particles in
%collision.
%
%Once the collision-resolution algorithm is implemented, we will check
%the computing performance first before we embark on incorporating
%fluctuating hydrodynamics~\cite{Bao17,Bao18}, which might be important
%even at the scales of the amphiphilic particle size in some cases. We
%propose to extend the high-order time-integration scheme for a
%stochastic differential equation~\cite{fu2015pre} to this system.
